Inferring the effective start dates of non-pharmaceutical interventions during \covid outbreaks

### 2020-07-08 ms/ms.tex
update WAIC table formatting; added RMSE table

We compare the posterior predictive plots of a model with a free $\tau$ with those of a model with~$\tau$ fixed at~$\tau^*$ and a model without~$\tau$ (i.e. transmission and reporting rates are constant).
All three models were fitted to case data up to Apr~11, used to predict out-of-sample case data up to Apr~23, and these predictions were then compared to the real case data. 
The model with free~$\tau$ clearly produces better and less variable predictions~(\autoref{fig:ppc}):
In all 11 of the European countries, the expected posterior RMSE (root mean squared error) of the out-of-sample predictions is lowest for the model with a free~$\tau$ (\autoref{table:RMSE}).
% TODO RMSE reference? (Ilia)
% TODO add table for RMSE (Ilia)
% TODO consider dropping WAIC for RMSE (both)
When we compare the models using WAIC~(\autoref{eq:WAIC}, \autoref{table:WAIC}), the model with a free~$\tau$ is strongly preferred in 9 out of 12 countries, the exceptions being Norway (only slight preference), Sweden, and Denmark. Notably, the data for Sweden and Denmark does not have a peak during the evaluated dates, possibly leading to wide 95\% confidence intervals on $\hat{tau}$ (\autoref{fig:tau-summary}) and poor WAIC in the model with free~$\tau$, whereas the duration between the first and last interventions was especially long in Norway (\autoref{table:NPI_dates}).

@@

%%% Early
\paragraph*{Early effective start of NPIs.}
In contrast, in some regions we estimate an effective start of NPIs $\hat{\tau}$ that is \emph{earlier} then the official date $\tau^*$ (\autoref{fig:tau-summary}).

Germany... %TODO (Yoav) \autoref{fig:early}

Sweden and Denmark show early effective start of NPIs. However, for both countries, the casa data does not include a "peak", that is, the number of daily cases continues to grow during the evaluated dates. We suspect this increases the uncertainty in our inference.
Accordingly, WAIC of our model with a free $\tau$ parameter was higher (less fit) than the WAIC of a model with $\tau$ fixed at $\tau^*$ the official date.

@@

% UPDATE
As countries relieve lockdowns and ease restrictions and the number of cases increases, we expect similar delays and advances to occur: in some countries the population will behave as if restrictions were eased even before the official date, and in some countries the population will continue to self-restrict even after restrictions are officially removed.

@@

% Table - WAIC %%% TODO regenerate table
% generated with /python Table-WAIC.py/
\begin{table}[h]
\centering
\pgfplotstabletypeset[
    col sep=comma,
    string type,
    every head row/.style={before row=\hline,after row=\hline},
    every last row/.style={after row=\hline},
    ]{../figures/Table-WAIC.csv}
\caption{
\textbf{WAIC values for the different models.}
	WAIC (widely applicable information criterion; \autoref{eq:WAIC})~\citep{gelman2013bayesian} values for models  with: no $\tau$ at all, \emph{No}; $\tau$ fixed at the official last NPI date $\tau^*$, \emph{Fixed}; and free parameter $\tau$, \emph{Free}. WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy and a difference of 2 is a popular significance level for model comparison~\citep{Kass1995}. Bold values emphasize cases in which the \emph{Free} model has the lowest WAIC.
}
\label{table:WAIC}
\end{table}

% Table - RMSE %%% TODO regenerate table
% generated with /python Table-RMSE.py/
%\begin{table}[h]
%\centering
%\pgfplotstabletypeset[
%    col sep=comma,
%    string type,
%    every head row/.style={before row=\hline,after row=\hline},
%    every last row/.style={after row=\hline},
%    ]{../figures/Table-RMSE.csv}
%\caption{
%\textbf{Posterior RMSE of out-of-sample predictions with the different models.}
%	Expected posterior predictive RMSE (root mean squared error) for models  with: no $\tau$ at all, \emph{No}; $\tau$ fixed at the official last NPI date $\tau^*$, \emph{Fixed}; and free parameter $\tau$, \emph{Free}.
%	In all cases, the model with free parameter $\tau$ has the lowest RMSE.
%	Models were fitted to case data up to Apr~11, 2020, and then used to generate 1,000 predictions from Apr~12 to Apr~24 by sampling model parameters from the posterior distribution.
%	These predictions were then compared to the real data using RMSE, and the mean RMSE value is shown in the table for each country and model.
%}
%\label{table:RMSE}
%\end{table}

>>>>

We compare the posterior predictive plots of a model with a free $\tau$ with those of a model with~$\tau$ fixed at~$\tau^*$ and a model without~$\tau$ (i.e. transmission and reporting rates are constant).
All three models were fitted to case data up to Apr~11, used to predict out-of-sample case data up to Apr~23, and these predictions were then compared to the real case data. 
The model with free~$\tau$ clearly produces better and less variable predictions~(\autoref{fig:ppc}):
In all 11 of the European countries, the expected posterior RMSE (root mean squared error) of the out-of-sample predictions is lowest for the model with a free~$\tau$ (\autoref{table:RMSE}).
When we compare the models using WAIC~(\autoref{eq:WAIC}, \autoref{table:WAIC}), the model with a free~$\tau$ is strongly preferred in 9 out of 12 countries, the exceptions being Norway (only slight preference), Sweden, and Denmark. Notably, the data for Sweden and Denmark does not have a peak during the evaluated dates, possibly leading to wide 95\% confidence intervals on $\hat{tau}$ (\autoref{fig:tau-summary}) and poor WAIC in the model with free~$\tau$, whereas the duration between the first and last interventions was especially long in Norway (\autoref{table:NPI_dates}).

@@

%%% Early
\paragraph*{Early effective start of NPIs.}
In some regions we estimate an effective start of NPIs $\hat{\tau}$ that is \emph{earlier} then the official date $\tau^*$ (\autoref{fig:tau-summary}).
The only significant early case is Germany, in which we estimate the effective start of NPIs at Mar~19 ($\pm$1.05 days 95 \%CI; \autoref{fig:early}).
This date occurs between the the first and last official NPI dates, which are Mar~12 and Mar~22 (\autoref{table:NPI_dates}). Therefore, when we refer to this case as "early", we mean that the effective date (Mar~19) occurs \emph{before} the official lockdown date (Mar~22), not that it occurs before all the NPIs.
Interestingly, Germany has the longest duration between first and last NPIs except in Norway (10 and 12 days, respectively; see \autoref{table:NPI_dates}).Nevertheless, the credible intervals for the effective date in Germany is very narrow ($\pm$1.05 days 95 \%CI; \autoref{fig:early}), whereas they are very wide in Norway (XXX).

We also estimate early effective start of NPIs in Sweden and Denmark. However, in both countries, the casa data does not include a "peak", that is, the number of daily cases continues to grow during the evaluated dates. We suspect this increases the uncertainty in our inference.
Accordingly, WAIC of our model with a free $\tau$ parameter was higher (less fit) than the WAIC of a model with $\tau$ fixed at $\tau^*$ the official date.

@@

As different countries experiment with various interventions strategies, we expect similar delays and advances to occur: in some cases the population will be late to comply with are guidelines, whereas in other cases the population will adopt restrictions or relaxations even before they are formally declared.

@@

% Table - WAIC %%% TODO regenerate table
% generated with /python Table-WAIC.py 2020-05-26-Apr11/
\begin{table}[h]
\centering
\pgfplotstabletypeset[
    col sep=comma,
    string type,
    every head row/.style={before row=\hline,after row=\hline},
    every last row/.style={after row=\hline},
    ]{../figures/Table-WAIC.csv}
\caption{
\textbf{WAIC values for the different models.}
	WAIC (widely applicable information criterion; \autoref{eq:WAIC})~\citep{gelman2013bayesian} values for models  with: no $\tau$ at all, \emph{No}; $\tau$ fixed at the official last NPI date $\tau^*$, \emph{Fixed}; and free parameter $\tau$, \emph{Free}. WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy and a difference of 2 is a popular significance level for model comparison~\citep{Kass1995}. Bold values emphasize cases in which the \emph{Free} model has the lowest WAIC.
}
\label{table:WAIC}
\end{table}

% Table - RMSE 
% generated with /python Table-RMSE.py 2020-05-26-Apr11/
\begin{table}[h]
\centering
\pgfplotstabletypeset[
    col sep=comma,
    string type,
    every head row/.style={before row=\hline,after row=\hline},
    every last row/.style={after row=\hline},
    ]{../figures/Table-RMSE.csv}
\caption{
\textbf{Posterior RMSE of out-of-sample predictions with the different models.}
	Expected posterior predictive RMSE (root mean squared error) for models  with: no $\tau$ at all, \emph{No}; $\tau$ fixed at the official last NPI date $\tau^*$, \emph{Fixed}; and free parameter $\tau$, \emph{Free}.
	In all cases, the model with free parameter $\tau$ has the lowest RMSE.
	Models were fitted to case data up to Apr~11, 2020, and then used to generate 1,000 predictions from Apr~12 to Apr~24 by sampling model parameters from the posterior distribution.
	These predictions were then compared to the real data using RMSE, and the mean RMSE value is shown in the table for each country and model.
}
\label{table:RMSE}
\end{table}
